Reciprocal Tolerance: Why States Allow Their Adversaries to Practice Espionage with Minimal Consequences

### 2021-05-26 article.tex
Write legal section

\maketitle
\newpage
\input{abstract.tex}
\newpage
\section{Introduction}
In 2015, federal authorities discovered a massive data breach inside the United States Office of Personnel Management (OPM). Hackers had, over the course of months, downloaded reams of sensitive data, including a cache of applications for federal security clearance. These applications exposed not just the personal information of the applicants themselves but also the detailed information they had supplied about their family and friends, including social security numbers, job applications, and home addresses---compromising the information of 22.1 million people in total.\footcite{nakashima_hacks_2015} US officials feared that the stolen data could help foreign intelligence services unmask undercover CIA officers, and the information could identify American officials who might be susceptible to blackmail, or worse, recruitment. FBI Director James Comey admitted that, with the kind of information leaked, a foreign intelligence service could ``know every place I've lived since I was 18, contact people at those addresses, neighbors at those addresses, all of my family, every place I've traveled outside the United States.''\footcite[Comey is referring specifically to his SF-86, the form that one files with OPM to apply for a security clearance.]{nakashima_hacks_2015} In the case of the OPM hack, it wasn't necessary to imagine who was in possession of that information; the US had already established that the Chinese military was likely behind the sophisticated attack.\footcite{koerner_inside_2016} But even though it possessed conclusive digital forensics, the Obama administration refused to name China as the culprit.\footnote{National Security Advisor John Bolton was the first American official to formally acknowledge that the Chinese were behind the OPM hack, three years later, in September 2018. See \cite{sanger_trump_2018}}

@@

Espionage is by far the most common form of cyberattack and governments do not always reveal when they have been compromised, so it is not terribly convincing to say that most cyber-espionage was ignored by the US Government while all 2 out 2 incidents of cyber-sabotage were not. But if we expand our scope globally for just a moment, it becomes apparent that the same pattern broadly holds. Stripped of all other context, espionage is the single least likely form of cyberattack to generate any type of public response.

@@

Based on this data, one could hypothesize that espionage does not always merit the same level of retaliation that sabotage or financial theft do because espionage is the least damaging form of cyberattack. While this makes sense in theory, it is not obviously true of the cyber-sabotage we've seen in practice. Consider how the OPM hack compares to a contemporary incident of state-sponsored sabotage. In 2014, hackers affiliated with the North Korean government leaked damaging emails from Sony executives and wiped some of Sony's servers.\footcite[This hack was allegedly in retaliation for the upcoming James Franco/Seth Rogen comedy about assassinating North Korean leader Kim Jong-Un.]{barnes_sony_2014} White House Press Secretary Josh Earnest alleged that the Sony hack was \enquote{destructive activity with malicious intent that was initiated by a sophisticated actor,} which he said merits \enquote{a proportional response.}\footcite{nakashima_white_2014} The Sony hack would be the first time a sitting US president publicly accused a foreign nation of launching a cyberattack and promised retaliation.\footcite{sanger_u.s._2016} The US made good on its threat and imposed sanctions on North Korea the following year.\footcite{lederman_us_2015} So why did the Sony hack---a newsworthy but far less widespread data breach---receive a significantly more serious response than a direct attack on a federal office that stole the private information of 22 million Americans?

@@

To the US government, stealing classified data from a federal office looks like something a spy might do, but wiping a movie studio's servers does not. While this explanation might ring true---cyber-espionage is espionage, sure---it is also essentially a tautology. It raises the question: why does it matter whether something qualifies as espionage, and is that a label that continues to apply even when the scale of what espionage can accomplish is magnified by new technology? Like a traditional intelligence operation, the OPM hack \enquote{merely} involved stealing sensitive information, but the quantity of compromised data was orders of magnitude beyond what could be accomplished by photographing the contents of a file cabinet and passing it along in a briefcase. \enquote{This is one of those cases where you have to ask, `Does the size of the operation change the nature of it?'\thinspace} said one anonymous senior intelligence official. \enquote{Clearly, it does.}\footcite{sanger_u.s._2015} The actions of the US government still say otherwise: at least for now, even operations on the scale of China's OPM infiltration do not merit significant retaliation.

This article seeks to explain how an old set of espionage norms continues to weather disruptive technological change. To do so first requires one to explain how it is possible that the routine minimization of espionage response happens at all. Extant espionage literature tends to focus on prescriptive questions---should espionage exist? is espionage legal?---rather than descriptive ones. There are few academic scholars who examine the theoretical implications of espionage in practice: what does it mean for existing theories of IR that a category of inter-state action routinely receives retaliation far less severe than the action's national security implications would seem to merit? In the first section of this paper I review the existing scholarship on norms and note the areas in which espionage does not neatly fit into the existing research.

In the latter section of this paper I demonstrate how espionage norms have been deployed during previous periods of tectonic technological change. The historical record demonstrates that the idea of espionage is deeply meaningful to the individual actors that decide state policy; if an action is considered to be espionage, it exists in a self-contained ontological space, with rules and consequences unaffected by rational cost analysis. The people who are in charge of these matters---what operations to undertake, how much risk to absorb, whether to respond to an operation conducted against them---intuit whether something \emph{counts as} espionage, and adjust their own actions to fit within that paradigm. The benefits for doing so are clear: if the aggrieved party considers an operation to be espionage, it is much less likely to retaliate, even when the consequences for that operation are deeply damaging. Cyber-espionage is a novel challenge for the intelligence community, but it is not the first paradigmatic shift in intelligence capabilities. Spy planes and spy satellites radically re-defined what espionage could accomplish and changed the course of the Cold War---the consequences for using them did not evolve accordingly.

\newpage

% This article seeks to unravel that tautology by first by explaining how states determine what constitutes \enquote{traditional} espionage and then by analyzing what strategic purpose that distinction serves.

@@

% Add information about my conclusions here

\subsection{Defining espionage norms}
There are few extant definitions of peacetime espionage, and no satisfying ones. The problem that aspiring espionage scholars inevitably encounter is that \enquote{peacetime espionage} is universally understood as a term of art but almost never defined by legal codes, or interpreted by case law. Most scholars who tackle this question settle on a definition like the one proposed by I\~{n}aki Navarrete and Russell Buchan: \enquote{\enquote{Peacetime espionage} is a colloquial rather than a legal term [used] to describe different methods of collecting confidential information from closed as opposed to open sources \textelp{} generally conducted by States, against States.}\footcite[901-902]{navarrete_out_2019} This definition purposefully excludes forms of extralegal government intervention that are often conflated with espionage, such as assassinations, election interference, and covert operations.\footnote{Often papers won't take a stand on defining espionage, but rather offer a working definition just to bound the scope of that particular paper. See \cite[600]{radsan_unresolved_2007}, \cite[2]{kapp_spying_2007})} Excluding these activities is important: assassinations and similar covert operations are not considered espionage, so they do not typically receive the protections of espionage norms.

For the purposes of this paper, I define espionage norms as the common understanding of how states are expected to respond to activities that remain within the bounds of espionage.\footnote{There are, of course, other \enquote{norms} that involve espionage. The so-called \enquote{Moscow Rules} dictate an intelligence officer are supposed to conduct themselves in hostile territory. In order to avoid constantly specifying, espionage norms are more strictly defined to include only the ones that affect IR.} Espionage norms prescribe that when one state uncovers an espionage operation conducted against it, the consequences imposed on the spying country remain minimal. Rarely does the response amount to more than a \emph{pro forma} denunciation of the activity and implementation of the necessary countermeasures to end it---the consequences for espionage are borne by the agents who commit it, not the state that sent them. In rare cases when additional costs \emph{are} imposed on a state for conducting a traditional espionage operation, they are not costs that would discourage the aggressor from attempting such an operation again.

The limits of espionage norms are defined by what intelligence professionals call \enquote{the rules of the game.} When states do impose harsher penalties, they often justify it by saying that they are punishing an operation that did not abide by the rules. The poisoning of Sergei Skripal, allegedly by Russian operatives, was widely described a violation of the rules.\footcite{masters_has_2018} Assassinations are off-limits, especially against former spies who have been traded to safety, and the US and the EU have issued multiple rounds of sanctions in response.\footcite{reuters_e.u._2019} Assassinations, however, are not the typical intelligence operation, and when an operation is understood by both parties to have been routine espionage, the consequences are minimized accordingly.

While most of the information that we have on espionage norms is qualitative, there is some data to support the limited consequences of espionage norms as well. Once again we can use cyber-espionage as an analogue for traditional espionage. Of the hundreds of state-sponsored cyberattacks known to the public, attacks classified as espionage are by far the most common. Cyber-espionage is, however, easily the least likely form of cyberattack to garner an official response from the attacked state

In the world of cyber-incidents, if it looks like espionage then it is espionage, and it will receive the customary diplomatic benefits associated with that characterization.\footnote{In programming, this is a principle called \enquote{Duck Typing.} If it looks like a duck, and it quacks like a duck, then it is a duck.}

@@

\subsection{Extant explanations for espionage norms}
Prior literature has attempted to explain espionage norms using international law, but in the absence of any significant treaties governing peacetime espionage, scholars are often left without definitive answers.\footnote{Because it is universally agreed that international law does not deal with peacetime espionage directly, many journal articles that deal with this question frame their work almost Socratically, devoting two sections to cases for and against the legality of espionage. In fact, the first major work on law and peacetime espionage is a series of commentaries collected in a single publication that respond to each other's arguments. See \cite{wright_essays_1962}} Espionage is universally illegal but every state acknowledges that allies and adversaries alike are guaranteed to engage in it.\footnote{Further complicating matters, espionage is the rare crime that might not be a crime depending on who the victim is. If an American government employee walks out of The Pentagon and hands classified documents to the KGB, they are at risk prosecuted as a traitor; if an American case officer hands classified KGB documents over to the CIA, they will be lauded as a hero. This example, in various forms, is frequently used to illustrate the impossibility of creating a consistent legal doctrine for espionage.} Because of that tension, scholars have typically treated espionage as \emph{sui generis}---starting from the assumption that espionage norms exist in a class of their own, then attempting to fit a legal doctrine around that fact. Jared Beim concludes that while peacetime espionage is often tolerated, it clearly constitutes an \enquote{intervention} as defined by the International Court of Justice, so there ought to be stronger enforcement mechanisms in place.\footnote{Beim also suggests that a weaker country could seek justice for an espionage operation by appealing to international organizations like the United Nations or the Council of Europe. For more, see \cite{beim_enforcing_2018}} Beim acknowledges, however, that at best \enquote{it is perhaps possible to stop the more egregious violations.}\footcite[672]{beim_enforcing_2018} A. John Radsan employs a more philosophical approach. Suggesting that we should resist the \enquote{Hegelian impulse} to resolve the tension between espionage and international law, Radsan concludes that \enquote{beyond any international consensus, countries will continue to perform espionage to serve their national interests \textelp{} International law does not change the reality of espionage.} Legal scholars of espionage generally reach the same conclusion: it would be almost impossible to argue that espionage is legal, but the international system leaves this tension in a state of salutary neglect. Taken to its logical endpoint, this branch of scholarship would simply argue that the espionage norms we observe today result from simple inertia.

@@

\section{Theory}
\subsection{The nuclear option}
% Add a poll about American knowledge of the CIA?

Civilian study of espionage is sometimes treated with skepticism because espionage agencies, by definition, operate with maximum possible secrecy. That secrecy does not inoculate these agencies from critical analysis. While the operations of a modern intelligence service are opaque, the decision to maintain one is not. American citizens, as well as their international counterparts, are broadly aware that their nations conduct extensive intelligence operations abroad. The Central Intelligence Agency (CIA) is funded by the United States Congress and reports directly to the President---it even promises the possibility of a career in espionage by offering internship positions for college students. Keeping these institutions intact is partly a function of inertia, but it is also an ongoing argument that espionage is a legitimate, even critical, component of statecraft.

Using traditional methods of analysis, espionage certainly seems like something that states should be doing more to discourage. The results of intelligence operations can have devastating, even existential effects on national security. In recent years China has relentlessly targeted American military secrets. A United States Navy report from March 2019 declared Chinese intelligence operations so extensive that they had substantively altered the balance of power between the two states.\footcite{lubold_navy_2019} \enquote{Long-term, US future military advantage is being diminished by years of IP exfiltration from [US military agencies], all with little to no adverse consequences to the thieves.}\footcite[6]{bayer_cybersecurity_2019} The forecasts were dire: \enquote{If the current trend continues unimpeded, the US will soon lose its status as the dominant global economic power.}\footcite[5]{bayer_cybersecurity_2019} The US Navy is not a disinterested party, of course, and we should be cautious about its conclusions; modern military technology is remarkably complex and China cannot simply steal its way to military dominance.\footnote{For a good discussion of these limitations, see: \cite{gilli_why_2019}} Nonetheless, espionage presents alarming opportunities for rival states to gain a comparative advantage and the consequences for taking advantage of them remain minimal.

% Since this generally doesn't happen between major world powers, one of two things must be true: either a state is miscalculating the relative gains it achieves through espionage, or the mutual tolerance of espionage has benefits that are sufficiently important to maintain even if one state gains less from it than its adversary.

% Certainly there are factors that the go into diplomatic decisions that the general public will likely never know. There are few reasons to believe, however, that this is more true for intelligence-related diplomatic matters than for generic ones. Every diplomatic decision is made with the backing of classified intelligence.

Not only are the risks associated with practicing espionage minimal, espionage norms limit the acceptable repercussions to ones that cannot possibly threaten the structure of civilian intelligence services. Were it more common to respond to espionage by imposing financial sanctions or even inflicting a military cost, a state would have to consider the possibility that attempting espionage could adversely impact other aspects of its security, leaving it worse off than when it started. States often need to make these calculations in other high-stakes arenas; pursuing a nuclear weapons program or protectionist trade policies could lead to severe economic retaliation. Instead, when an especially damaging espionage operation is uncovered, the traditional punishment is to declare certain embassy officials \emph{persona non grata}, ejecting them from the country. This action often results in an equal and opposite retaliation: \enquote{tit-for-tat} diplomatic expulsions. Both states will expel a portion of the other's embassy staff that they suspect of being spies. Losing intelligence officers in an embassy temporarily makes it more difficult to conduct espionage, but it doesn't make a state less likely to try. According to John Sipher, former head of the CIA's Russia operations, previous Western expulsions have never crippled Russian intelligence collection. \enquote{If you have 150-200 intelligence officers in-country, losing 50 is painful, but hardly debilitating.} \footcite{dettmer_united_2018}

>>>>

\maketitle
\newpage
\input{abstract.tex}
\newpage
\section{Introduction}
In 2015, federal authorities discovered a massive data breach inside the United States Office of Personnel Management (OPM). Hackers had, over the course of months, downloaded reams of sensitive data, including a cache of applications for federal security clearance. These applications exposed not just the personal information of the applicants themselves but also the detailed information they had supplied about their family and friends, including social security numbers, job applications, and home addresses---compromising the information of 22.1 million people in total.\footcite{nakashima_hacks_2015} US officials feared that the stolen data could help foreign intelligence services unmask undercover CIA officers, and the information could identify American officials who might be susceptible to blackmail, or worse, recruitment. FBI Director James Comey admitted that, with the kind of information leaked, a foreign intelligence service could ``know every place I've lived since I was 18, contact people at those addresses, neighbors at those addresses, all of my family, every place I've traveled outside the United States.''\footnote{\cite{nakashima_hacks_2015}. Comey is referring specifically to his SF-86, the form that one files with OPM to apply for a security clearance.} In the case of the OPM hack, it wasn't necessary to imagine who was in possession of that information; the US had already established that the Chinese military was likely behind the sophisticated attack.\footcite{koerner_inside_2016} But even though it possessed conclusive digital forensics, the Obama administration refused to name China as the culprit.\footnote{National Security Advisor John Bolton was the first American official to formally acknowledge that the Chinese were behind the OPM hack, three years later, in September 2018. See \cite{sanger_trump_2018}}

@@

Espionage is by far the most common form of cyberattack and governments do not always reveal when they have been compromised, so it is not terribly convincing to say that most cyber-espionage was ignored by the US Government while both of the two known incidents of cyber-sabotage were not. After all, each cyber-incident exists in the context of the American relationship with the nation that perpetrated it; the decision to respond or retaliate is influenced by countless factors and there simply are not enough incidents of cyber-sabotage against the United States to establish a clear contrast in the degree of retaliation. But if we expand our scope globally for just a moment, it becomes apparent that the same pattern broadly holds. Stripped of all other context, espionage is the single least likely form of cyberattack to generate any type of public response.

@@

Based on this data, one could hypothesize that espionage does not always merit the same level of retaliation that sabotage or financial theft do because espionage is the least damaging form of cyberattack. While this makes sense in theory, it is not obviously true of the cyber-sabotage we've seen in practice. Consider how the OPM hack compares to a contemporary incident of state-sponsored sabotage. In 2014, hackers affiliated with the North Korean government leaked damaging emails from Sony executives and wiped some of Sony's servers.\footnote{\cite{barnes_sony_2014}. The Sony hack was allegedly in retaliation for the upcoming James Franco/Seth Rogen comedy about assassinating North Korean leader Kim Jong-Un.} White House Press Secretary Josh Earnest alleged that the Sony hack was \enquote{destructive activity with malicious intent that was initiated by a sophisticated actor,} which he said merits \enquote{a proportional response.}\footcite{nakashima_white_2014} The Sony hack would be the first time a sitting US president publicly accused a foreign nation of launching a cyberattack and promised retaliation.\footcite{sanger_u.s._2016} The US made good on its threat and imposed sanctions on North Korea the following year.\footcite{lederman_us_2015} So why did the Sony hack---a newsworthy but far less widespread data breach---receive a significantly more serious response than a direct attack on a federal office that stole the private information of 22 million Americans?

@@

According to the US government, stealing classified data from a federal office looks like something a spy might do, but wiping a movie studio's servers does not. While this explanation might ring true---cyber-espionage is espionage, sure---it is also essentially a tautology. It raises the question: why does it matter whether something qualifies as espionage, and is that a label that continues to apply even when the scale of what espionage can accomplish is magnified by new technology? Like a traditional intelligence operation, the OPM hack \enquote{merely} involved stealing sensitive information, but the quantity of compromised data was orders of magnitude beyond what could be accomplished by photographing the contents of a file cabinet and passing it along in a briefcase. Said one anonymous senior intelligence official: \enquote{This is one of those cases where you have to ask, \enquote{Does the size of the operation change the nature of it?} Clearly, it does.}\footcite{sanger_u.s._2015} The actions of the US government would ultimately say otherwise; at least for now, even operations on the scale of China's OPM infiltration do not merit significant retaliation.

This article seeks to explain how an old set of espionage norms continues to weather disruptive technological change. To do so first requires one to explain how it is possible that the routine minimization of espionage response happens at all. Extant espionage literature tends to focus on prescriptive questions---should espionage exist? is espionage legal?---rather than descriptive ones. There are few academic scholars who examine the theoretical implications of espionage as it exists in practice: what does it mean for existing theories of IR that a category of inter-state action routinely receives retaliation far less severe than the action's national security implications would seem to merit? In the first section of this paper I review the existing scholarship on norms and note the areas in which espionage does not neatly fit into the existing research.

In the latter section of this paper I demonstrate how espionage norms have been deployed during previous periods of tectonic technological change. The historical record demonstrates that the idea of espionage is deeply meaningful to the individual actors that decide state policy; if an action is considered to be espionage, it exists in a self-contained ontological space, with rules and consequences that operate outside rational cost analysis. The people who are in charge of these matters---what operations to undertake, how much risk to absorb, whether to respond to an operation conducted against them---intuit whether something \emph{counts as} espionage, and adjust their own actions to fit within that paradigm. The benefits for doing so are clear: if the aggrieved party considers an operation to be espionage, it is much less likely to retaliate, even when the consequences for that operation are deeply damaging. Cyber-espionage is a novel challenge for the intelligence community, but it is not the first paradigmatic shift in intelligence capabilities. Spy planes and spy satellites radically re-defined what espionage could accomplish and changed the course of the Cold War---the consequences for using them did not evolve accordingly.

\section{Possible Theories}
Few nations formally acknowledge that peacetime espionage is a core function of the state, so you rarely see treaties or statutes properly define it. The Espionage Act of 1917 implicitly defines espionage as providing \enquote{information relating to the national defense which information the possessor has reason to believe could be used to the injury of the United States or to the advantage of any foreign nation [\textellipsis] to any person not entitled to receive it.}\footnote{\cite{noauthor_notitle_nodate}. I say \enquote{implicitly} because the Espionage Act of 1917 never actually defines espionage, it's just a piece of legislation called \enquote{The Espionage Act} that criminializes the quoted behavior.} Passed right after US entry into World War I, The Espionage Act was written to address wartime espionage (each clause contains an exhaustive list of military installations, from torpedo stations to canals, within which one might purloin damaging information) but remarkably remains the sole legal tool for prosecuting peacetime espionage as well; almost a century later, the United States charged Edward Snowden with two counts brought under the Espionage Act for providing information about domestic surveillance programs to the press.\footnote{\cite{finn_us_2013}. Even though the text of the law does not specify that charges can only be levied during wartime, the Espionage Act precedes the modern concept of peacetime espionage by decades. The CIA was founded in 1497, and its main predecessor, the military Office of Strategic Services, did not exist prior to the Second World War.} Procedures governing the appropriate practice of wartime espionage are outlined in both the Geneva Convention and its precursor, the Hague Conventions of 1907, but the legal status of peacetime espionage remains likewise unresolved in international law.\footcite[651]{beim_enforcing_2018}

The definition of espionage that best captures the phenomenon discussed here is the one proposed by I\~{n}aki Navarrete and Russell Buchan: \enquote{\enquote{Peacetime espionage} is a colloquial rather than a legal term [used] to describe different methods of collecting confidential information from closed as opposed to open sources \textelp{} generally conducted by States, against States.}\footcite[901-902]{navarrete_out_2019} This definition purposefully excludes covert operations that are often conflated with espionage, such as assassinations, election interference, and other forms of extralegal government intervention.\footnote{Some scholars also choose not to export their own definition of espionage, but rather offer a working definition just to bound the scope of that particular paper. See \cite[600]{radsan_unresolved_2007}; \cite[2]{kapp_spying_2007}} Covert operations are often conducted by intelligence agencies and by people who would call themselves spies, but do not universally receive the \enquote{you gotta hand it to them} light touch that more strictly-defined espionage does.\footnote{This article excludes covert action from espionage because covert action results in diplomatic consequences, but the question of whether \enquote{intelligence} includes covert action is disputed. While I employ the Navarrete and Buchan definition, for a contrasting definition that includes covert action, see: \cite[518]{stout_intelligence_2018}.} When Russian operatives were alleged to have poisoned Sergei Skripal, this was widely to considered to have violated \enquote{the rules of the game,} so the US and the EU issued multiple rounds of sanctions in response.\footnote{\cite{masters_has_2018}. Not only are assassinations considered in violation of the rules, but Skripal was a former British spy who had been traded back to the UK in exchange for a Soviet spy, which is especially off-limits.} Espionage is far more common than assassinations however, and when an operation is understood by both parties to have been classic espionage, both parties routinely minimize the consequences associated with it.

% Rarely does the response amount to more than a \emph{pro forma} denunciation of the activity and implementation of the necessary countermeasures to end it---the consequences for espionage are borne by the agents who commit it, not the state that sent them.

To attribute state decisions of that magnitude to violations of \enquote{the rules of the game} is painfully colloquial, but as Navarrete and Buchan smartly identify, the very business of peacetime espionage is colloquial. The challenge an espionage scholar faces is to impose some orderly framework of analysis on a field that lacks the key structures---laws, treaties, promises, wars---traditionally used to build such frameworks. The remainder of this section will explore whether existing theories of international relations can explain the disparity between the harm that espionage can accomplish and the reluctance of states to impose the sorts of consequences (economic or military) that would discourage its use.\footnote{If at this point in time you are not sufficiently convinced that espionage \emph{is} something that states treat with a different set of consequences, please be aware that the evidentiary section of this article will examine cases in which this disparity is especially stark.}

@@

@@

\subsection{The curious case of espionage norms}
The informal yet widely consistent set of espionage consequences is consistent with the manner in which international relations (IR) norms operate. The canonical framework for understanding how IR norms are created and promoted is the one outlined by Martha Finnemore and Kathryn Sikkink, who define a norm \enquote{as a standard of appropriate behavior for actors with a given identity.}\footcite[891]{finnemore_international_1998} In this case, espionage norms govern the standard of appropriate behavior for actors who wish to be considered spies; if said actor's behavior conforms to those standards, then the consequences for their actions will be minimized accordingly. They also note that because norms embody a sense of \enquote{oughtness,} norms are employed when justification for some action is required, and therefore \enquote{leave an extensive trail of communication among actors that we can study.}\footcite[892]{finnemore_international_1998} The paper trail that the Eisenhower spy program left behind---and the justifications that administration made for using its various devices---is the evidentiary focus of this article.

What sets espionage apart from the typical IR norm is that there is little evidence of anyone arguing that espionage \emph{should} work in any particular way, only that it \emph{does} and therefore it makes sense to operate within those confines. Finnemore and Sikkink illustrate their theory of norm entrepreneurship using the women's rights movement and the development of the laws of war, both of which compare rather oddly to the norms of espionage.\footcite[894]{finnemore_international_1998} One struggles to imagine pro-espionage demonstrations analogous to the suffragette marches, and as we have already seen, peacetime espionage is entirely absent from the Geneva Convention and its antecedents. If norms embody \enquote{oughtness,} who is arguing that espionage ought to exist?

\subsection{Espionage in international law1}

One place you might look for evidence of norm entrepreneurship is international law. Norms are sometimes understood as an alternative to legal regimes, but laws and norms frequently reinforce each other: laws can establish new norms, or they can derive their legitimacy from preexisting norms.\footcite[441]{finnemore_constructing_2016} As Faesen et al. note, \enquote{one common goal of norm entrepreneurs is to construct ties to law that strengthen and refine the behavioral expectations of a given norm}\footcite[25]{faesen_blurred_2020} Therefore, if espionage norms exist, one might expect to find evidence that states have attempted to enshrine this set of norms into law, or referenced an existing law to defend their decision to treat an particular incident pursuant to espionage norms.

Instead, legal scholarship has been divided for 80 years on whether international law takes a position on peacetime espionage, and, if so, whether that position endorses or prohibits it. The inaugural text on the subject, \emph{Essays on Espionage and International Law,} presents the views of two opposing law professors, Julius Stone and Quincy Wright, both of whom are widely cited in the subsequent literature.\footnote{These two essays are primarily concerned with addressing the implications of the U-2 Incident, a new form of peacetime espionage. The facts of that incident will be discussed in-depth in the Case Studies section of this article.} Wright argues that \enquote{[a]ny act by an agent of one state committed in another state’s territory, contrary to the laws of the latter, constitutes intervention.}\footcite[13]{wright_espionage_1962} In his view, if espionage were not legally wrong, then governments would not disclaim their spies.\footcite[17]{wright_espionage_1962} Stone proclaims his \enquote{almost complete dissent} from Wright's view, noting that when spies are caught, the aggrieved state never attempts \enquote{to shift responsibility for the espionage to the home state.}\footcite[33]{stone_legal_1962} If peacetime espionage were a delinquency, Stone argues, why would states continue to allow the transparent excuses of their adversaries to stand?\footcite[39]{stone_legal_1962}

Modern literature on espionage largely builds on one of these two conclusions---either state tolerance of espionage is evidence for its legality, or state disavowal is evidence of its illegality. Taking after Wright, Jared Beim argues that while peacetime espionage is often tolerated, it clearly constitutes an \enquote{intervention} as defined by the International Court of Justice, so there ought to be stronger enforcement mechanisms in place.\footnote{\cite[672]{beim_enforcing_2018}. Beim suggests that a weaker country could seek justice for an espionage operation by appealing to international organizations like the United Nations or the Council of Europe. He acknowledges, however, that at best \enquote{it is perhaps possible to stop the more egregious violations.}} Jeffrey Smith argues the opposite, that because espionage is so widespread, \enquote{it is fair to say that the practice of states recognizes espionage as a legitimate function of the state, and therefore it is legal as a matter of customary international law.}\footcite[544]{smith_keynote_2007} He notes that the routine practice of declaring suspected intelligence officers \emph{persona non grata} legitimizes the practice of espionage as one that merely violates diplomatic protocol, not international law.\footcite[544]{smith_keynote_2007} A. John Radsan avoids taking a position in either direction, arguing that we should resist the \enquote{Hegelian impulse} to resolve the tension between espionage and international law at all. He concludes that \enquote{beyond any international consensus, countries will continue to perform espionage to serve their national interests \textelp{} International law does not change the reality of espionage.}\footcite[623]{radsan_unresolved_2007} Radsan's non-position is, of course, a position of its own; citing Radsan, Craig Forcese observes that \enquote{the very fact that there are three camps with such diametric positions itself suggests that \textelp{} there is no clear answer on the international legality of extraterritorial espionage.}\footcite[204-205]{forcese_spies_2011}

As the Forcese quote suggests, what unites the literature (and meta-literature) on peacetime espionage is agreement that, regardless of the author's opinion on whether this is desirable, nations have left espionage jurisprudence in a state of salutary neglect. Forcese writes that \enquote{the international community seems content with an artful ambiguity on the question.}\footcite[205]{forcese_spies_2011} Darien Pun arrives at the same conclusion, noting that despite the years of debate reviewed here, the status of peacetime espionage has remained unclear.\footcite[260]{pun_rethinking_2017} Though this article is not interested in the legality of espionage, it is extraordinary to note that legal scholars have been completely unable to determine it, because no state has even attempted to resolve this question. At least within the context of international law, there is no evidence of any norm entrepreneurship for peacetime espionage whatsoever.

\subsection{Espionage a framing device}
Not only is peacetime espionage a practically unquestioned normative regime, the proximity of an action to espionage is a first principle against which other norms are judged. Two key strategies that norm entrepreneurs employ are \emph{framing}, contextualizing the facts of a case in the light most favorable to promoting a desired norm, and \emph{linking}, connecting the desired norm to an existing, more powerful one.\footcite[451]{finnemore_constructing_2016} While the boundary between these two concepts is blurry, policymakers characterize espionage incidents in a manner much closer to the former. The OPM hack was framed as an \enquote{espionage issue,} the way someone might frame landmines as a \enquote{war crimes issue} or women's suffrage as a \enquote{human rights issue,} invoked as a fundamental principle whose general desirability is assumed.\footnote{Obviously, the notion that war crimes should be discouraged is itself a norm. The distinction that I believe the authors of this concept intended is that one frames a prospective norm in a manner where its desirability does not need to be explained, whereas one links to an existing norm to borrow that norm's explanation for its utility.}

The inverse framing is true as well: a norm that something is \emph{not} permissible can be pushed with the justification that it is \emph{not} espionage. The United States has long acted as a norm entrepreneur against economic espionage, the theft of commercial trade secrets for a nation's economic advantage, differentiating this particular form of espionage from the implied acceptability of traditional espionage.\footcite[19]{nye_normative_2018} The norm against economic espionage, as Samuel Rascoff describes it, \enquote{curiously gerrymandered} to forbid this form of market-state collaboration, while allowing other, functionally identical, ones.\footcite[250]{rascoff_norm_2016}

% what's less clear is what keeps that norm in place

% law is one place where it might be acknowledged

% talk about institutionalization, finnemore + sikkink 901

% Further complicating matters, espionage is the rare crime that might not be a crime depending on who the victim is. If an American government employee walks out of The Pentagon and hands classified documents to the KGB, they are at risk prosecuted as a traitor; if an American case officer hands classified KGB documents over to the CIA, they will be lauded as a hero. This example, in various forms, is frequently used to illustrate the impossibility of creating a consistent legal doctrine for espionage.

@@

% now do cybersec norms

% \subsection{The nuclear option}
% % Add a poll about American knowledge of the CIA?

% Civilian study of espionage is sometimes treated with skepticism because espionage agencies, by definition, operate with maximum possible secrecy. That secrecy does not inoculate these agencies from critical analysis. While the operations of a modern intelligence service are opaque, the decision to maintain one is not. American citizens, as well as their international counterparts, are broadly aware that their nations conduct extensive intelligence operations abroad. The Central Intelligence Agency (CIA) is funded by the United States Congress and reports directly to the President---it even promises the possibility of a career in espionage by offering internship positions for college students. Keeping these institutions intact is partly a function of inertia, but it is also an ongoing argument that espionage is a legitimate, even critical, component of statecraft.

% Using traditional methods of analysis, espionage certainly seems like something that states should be doing more to discourage. The results of intelligence operations can have devastating, even existential effects on national security. In recent years China has relentlessly targeted American military secrets. A United States Navy report from March 2019 declared Chinese intelligence operations so extensive that they had substantively altered the balance of power between the two states.\footcite{lubold_navy_2019} \enquote{Long-term, US future military advantage is being diminished by years of IP exfiltration from [US military agencies], all with little to no adverse consequences to the thieves.}\footcite[6]{bayer_cybersecurity_2019} The forecasts were dire: \enquote{If the current trend continues unimpeded, the US will soon lose its status as the dominant global economic power.}\footcite[5]{bayer_cybersecurity_2019} The US Navy is not a disinterested party, of course, and we should be cautious about its conclusions; modern military technology is remarkably complex and China cannot simply steal its way to military dominance.\footnote{For a good discussion of these limitations, see: \cite{gilli_why_2019}} Nonetheless, espionage presents alarming opportunities for rival states to gain a comparative advantage and the consequences for taking advantage of them remain minimal.

% % Since this generally doesn't happen between major world powers, one of two things must be true: either a state is miscalculating the relative gains it achieves through espionage, or the mutual tolerance of espionage has benefits that are sufficiently important to maintain even if one state gains less from it than its adversary.

% % Certainly there are factors that the go into diplomatic decisions that the general public will likely never know. There are few reasons to believe, however, that this is more true for intelligence-related diplomatic matters than for generic ones. Every diplomatic decision is made with the backing of classified intelligence.

% Not only are the risks associated with practicing espionage minimal, espionage norms limit the acceptable repercussions to ones that cannot possibly threaten the structure of civilian intelligence services. Were it more common to respond to espionage by imposing financial sanctions or even inflicting a military cost, a state would have to consider the possibility that attempting espionage could adversely impact other aspects of its security, leaving it worse off than when it started. States often need to make these calculations in other high-stakes arenas; pursuing a nuclear weapons program or protectionist trade policies could lead to severe economic retaliation. Instead, when an especially damaging espionage operation is uncovered, the traditional punishment is to declare certain embassy officials \emph{persona non grata}, ejecting them from the country. This action often results in an equal and opposite retaliation: \enquote{tit-for-tat} diplomatic expulsions. Both states will expel a portion of the other's embassy staff that they suspect of being spies. Losing intelligence officers in an embassy temporarily makes it more difficult to conduct espionage, but it doesn't make a state less likely to try. According to John Sipher, former head of the CIA's Russia operations, previous Western expulsions have never crippled Russian intelligence collection. \enquote{If you have 150-200 intelligence officers in-country, losing 50 is painful, but hardly debilitating.} \footcite{dettmer_united_2018}

% If removing a quarter of the officers doesn't do lasting damage, then why not remove the full 200? The swift expulsion of all suspected intelligence officers from a country---and the inevitable removal of the country's own intelligence officers in retaliation---would deeply injure both their espionage capabilities indefinitely.\footnote{Nothing on this scale has ever been attempted, but Operation Foot, the largest diplomatic expulsion in the Cold War, expelled 105 Soviet diplomats from the UK. While still not the total shutdown proposed in this section, Operation Foot \enquote{was a blow from which the Soviet intelligence effort in the UK never recovered.} For more, see: \cite{hughes_giving_2006-1}} This nuclear option is always available to both parties and yet, in modern history, it has never been exercised. We can reasonably infer that each nation's leadership is making an identical calculation: the benefits of continuing to practice espionage in my adversary's backyard outweigh the risks associated with allowing my adversary to do the same to me. The question remains how they can both be correct.
